---
title: "Exploring World Bank Data"
author: "Maria Benavides"
output: github_document
---

## Load necessary libraries

```{r}
library(tidyverse)
```

## Write a function to import the data files

```{r}

# Create the function that imports and tiddies the data frame

import_tiddy_data <- function(x) {
  country_data <- read_csv(x, skip = 3)  %>%
      rename(
        indicator_name = `Indicator Name`
      ) %>%
      select(
        -c(4:44),  # I decided to drop most years, to have a more simple data frame
        -X65
      ) %>%
      filter( # I added these variables thinking of GDP as a measure of development
        indicator_name %in% c("GDP per capita growth (annual %)", 
                         "Physicians (per 1,000 people)",
                         "Hospital beds (per 1,000 people)",
                         "Literacy rate, adult total (% of people ages 15 and above)", 
                         "Intentional homicides (per 100,000 people)", 
                         "Human capital index (HCI) (scale 0-1)", 
                         "Fertility rate, total (births per woman)", 
                         "Prevalence of undernourishment (% of population)", 
                         "Unemployment, total (% of total labor force) (national estimate)", 
                         "CO2 emissions (kg per PPP $ of GDP)", 
                         "Access to electricity (% of population)",
                         "GDP per capita, PPP (current international $)"                         
                         )
      ) %>%
      pivot_longer(
        cols = c(4:23), 
        names_to = "year", 
        values_to = "indicator_value"
      ) %>%
      mutate(
        year = as.numeric(year) 
      ) %>%
      pivot_wider(
        names_from = indicator_name, 
        values_from = indicator_value
      ) %>%
      rename(
        country = `Country Name`, 
        country_code = `Country Code`,
        gdp_growth = `GDP per capita growth (annual %)`, 
        gdp_pc = `GDP per capita, PPP (current international $)`, 
        physicians = `Physicians (per 1,000 people)`, 
        hospital_beds = `Hospital beds (per 1,000 people)`,
        literacy_rate = `Literacy rate, adult total (% of people ages 15 and above)`,
        homicide_1000ppl = `Intentional homicides (per 100,000 people)`, 
        hci = `Human capital index (HCI) (scale 0-1)`,
        fertility_rate = `Fertility rate, total (births per woman)`,
        undernourushment = `Prevalence of undernourishment (% of population)`,
        unemployment = `Unemployment, total (% of total labor force) (national estimate)`,
        co2_emissions = `CO2 emissions (kg per PPP $ of GDP)`,
        electricity = `Access to electricity (% of population)`
      )
    
 return(country_data) # Recall to return the desire output 
}    
  

```

## Import the data

```{r message=FALSE, warning=FALSE}

# Import all files using the function

countries <- list.files("data_world_bank")  # Create a vector with the names of all files

wb_database <- import_tiddy_data(
  paste(
    "data_world_bank/",countries[1], sep = "") # To make it more precise, use paste to merge 
                                               # folder name and files names
  )

for (country in countries[2:217]) {           # Use for loop to import all files at once
  country_database <- import_tiddy_data(
    paste(
      "data_world_bank/",country, sep = "")
    )
  wb_database <- rbind(wb_database, country_database) # Combine all data frames into one
}

```

## Explore the data

```{r}

theme_set(new = theme_minimal()) # I think I finally figured out the theme set issue 

# ANALYSIS 1 
# Calculate average growth GDP per capita for each country (2000-2019) and the average number 
# of physicians

average_gdp_phy <- group_by(wb_database, country) %>%
  summarise(
    average_gdp = mean(gdp_pc, na.rm = TRUE), 
    average_phy = mean(physicians, na.rm = TRUE)
  )

average_gdp_phy_nooutliers <- group_by(wb_database, country) %>% # Dropping the highest GDP
  summarise(
    average_gdp = mean(gdp_pc, na.rm = TRUE), 
    average_phy = mean(physicians, na.rm = TRUE)
  ) %>%
  filter(
    average_gdp <= 70000
  )

# Use a plot to determine is there is any relationship between the average GDP growth and average
# number of physicians per 1,000 people

ggplot(average_gdp_phy, aes(x=average_gdp, y=average_phy)) + 
  geom_point(size = 0.7) +
  labs(
    title="Relationship between GDP per capita and number of physicians", 
    x="Average GDP per capita growth", 
    y="Average number of physicians per 1,000 ppl"
    ) +
  geom_smooth(se = FALSE)


ggplot(average_gdp_phy_nooutliers, aes(x=average_gdp, y=average_phy)) + 
  geom_point(size = 0.7) +
  labs(
    title="Relationship between GDP per capita and number of physicians - no outliers", 
    x="Average GDP per capita growth", 
    y="Average number of physicians per 1,000 ppl"
    ) +
  geom_smooth(se = FALSE)

```

According to the first plot, there is **no clear relationship** between the average GDP per capita, and the annual average number of physicians per 1,000 people, considering data from 2000 to 2019.
Nonetheless, when we drop the outliers values (GDP > 70000 PPP), a more positive relationship starts to appear. 
In consideration of current events (COVID-19 pandemic) we could think that the most wealthy countries are better prepare to face this situation. 

```{r}

# ANALYSIS 2 
# Filter dataframe to use only the variable of interest and a given year 

co2_2014 <- wb_database %>%
  filter(
    year == 2014 # 2014 is the most recent year with more values for the CO2 variable
  ) %>% 
  select(
    country, 
    co2_emissions, 
    gdp_pc
  )
  
# Create a histogram to understand the distribution of the C02 emissions variable 

ggplot(co2_2014, aes(x=co2_emissions)) + 
  geom_histogram(binwidth = 0.1, color = "black", fill = "white") + 
  labs(
    title="Histogram of CO2 emissions during 2014", 
    x="CO2 emissions (kg per PPP $ of GDP)", 
    y="Count"
    )
  
# Find the relationship between CO2 emissions and GDP growth

ggplot(wb_database, aes(x=co2_emissions, y=gdp_pc)) + 
  geom_point(size = 0.7) +
  labs(
    title="Relationship between GDP per capita and CO2 emissions (2014)", 
    x="CO2 emissions (kg per PPP $ of GDP)", 
    y="GDP per capita growth (2014)"
    ) +
  geom_smooth(se = FALSE)

```

The histogram shows that most countries have a similar level of CO2 emissions. Nonetheless, there are some  exceptions. We could think, first, that those countries that produce more emissions, are the wealthiest, but the second graph shows no clear relationship between the two variables. 

```{r}

# ANALYSIS 3
# Filter dataframe to use only the variable of interest 

violence_col_mex <- wb_database %>%
  filter(
    country_code == "COL" | 
      country_code == "MEX" 
  ) %>% 
  select(
    country, 
    homicide_1000ppl, 
    year
  )
  
# Create a line graph to see the trend on homicides in Colombia and Mexico 

ggplot(violence_col_mex, aes(x=year, y=homicide_1000ppl)) + 
  geom_line(aes(color = country,), 
            linetype = "dashed") +
  scale_x_continuous(limits = c(2000, 2018),
                     breaks = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014, 2016)) + 
  geom_point(size=0.7) +
  labs(
    title="Intentional homicides (per 100,000 people) in Mexico and Colombia", 
    x="Year", 
    y="Intentional homicides (per 100,000 people)"
    )


```

Colombia and Mexico have suffered from internal conflicts and violence for decades. We can see in this graph, thought, that Colombia's situation is historically worse. Even after a peace agreement signed in 2016, Colombia had higher homicide rates. Mexico's situation is worsening, especially after the Calderon administration (2016). 

## Session info

```{r}
devtools::session_info()
```


